The Evolution of Phenotypically Invariant Genetic Networks

### 2017-12-01 kalman_walk_paper.tex
general edits to paper. Changed figure 6 (f2 vs f1 divergence)

A complex molecular machinery translates an organism's genome into the characteristics on which natural selection acts,
the phenotype.
% Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. 
%Attaining a general understanding of the functioning and evolution of this molecular machinery
It is an overarching goal of many biological subdisciplines to attain a general understanding of the function and evolution of this molecular machinery. 
For example, 
there is a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks
\citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}. 
Mathematical models of both particular regulatory networks
and the evolution of such systems in general
can provide guidance where intuition fails,
and thus has the potential to discover general principles in the organization of biological systems 
as well as provide concrete numerical predictions \citep{servedio2014not}.

@@

Although nonidentifiability may frustrate the occasional engineer or scientist, viewed from another angle,
this concept can provide a starting point for thinking about externally equivalent systems
-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
These functional symmetries manifest in convergent and parallel evolution, 
as well as \emph{developmental system drift}: the observation that
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
 % \jss{should hare2008sepsid be here? It's sequence but not ``system'' drift}
% are these good here? {taylor2016diverse -- maybe not, matsui2015regulatory -- I think this relevant, stergachis2014conservation}.
%
%AND THESE?
  %% Over the last several years, different computational models have been applied to study reproductive incompatibility and speciation in models of gene regulation \citep{porter2002speciation, tulchinsky, palmer2009dynamics, tulchinsky2014hybrid, khatari2015simple}.
  %\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model. Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybrid incompatibility when allopatric populations employ divergent readout languages. This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene. Furthermore, they report reproductive isolation primarily following directional selection for a change in expression levels in each allopatric population; the evidence for reproductive isolation following balancing selection is much weaker. Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection. 
%Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective. Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially. Otherwise hybrids between two allopatric populations have fairly high fitnesses.

In this paper we outline a theoretical framework to study the evolution of biological systems, such as gene regulatory networks.
We study the evolution of an optimally adapted population subject to stabilizing selection for phenotype -- that is the population evolves under constant selective and environmental pressures.
Even if the phenotype remains stable over evolutionary time, the underlying mechanism might not remain so,
%Even when stabilizing selection preserves phenotype over evolutionary time, the underlying mechanism may change, 
as many distinct (and mutationally connected) molecular pathways can realize identical phenotypes.
%where the optimal phenotype remains constant over evolutionary time,
%so that despite strong stabilizing selection for phenotype,
%neutral drift in the underlying genotype is possible.
Applying results from system theory, we present
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes, and show that any biological system, in principal, can undergo system drift, as there is never a unique system architecture per phenotype.
%which gives concrete expectations for how any such
%system can, in principal, undergo system drift and rewire.
Even under stabilizing selection, a population can explore the set of all possible phenotypically equivalent gene networks.
Phenotypically equivalent gene networks are not necessarily compatible with one another, as such system drift may result in the reproductive incompatibility between populations isolated for a sufficiently long period of time, even in the absence of any sort of adaptative, selective, or environmental change. Finally, in some cases we estimate that reproductive incompatibility due to system drift can manifest on timescales on the order of $N_e$ generations.

analogous to those in the Dobzhansky-Muller model. 
%Consequentially, two populations isolated for a sufficiently long period of time, 
%will likely produce inviable hybrids, despite the absence of adaptation, directional selection, or environmental change.
% Under these conditions, speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.

@@

Above we synthesize concepts and tools from molecular and evolutionary biology with tools from control theory
to study the evolution of a mechanistic model of the molecular genotype-phenotype map under stabilizing selection.
This model allows us to analytically describe the processes discussed verbally as
phenogenetic drift \citep{weiss2000phenogenetic} and developmental system drift \citep{true2001developmental}.
In this context, the Kalman decomposition \citep{kalman1963mathematical} implies that nearly all systems are nonidentifiable,
and gives an analytical description of all phenotypically equivalent gene networks.
System nonidentifiability, rather than being a computational nuisance,
implies the existence of axes of genetic variation that are not constrained by selection.
The independent movement of separated populations along these axes
can lead to reduction in hybrid viability, and hence speciation,
at a speed that depends on effective population size and amount of genetic variation. 
In this model, at biologically reasonable parameter values,
system drift is a significant -- and often rapid -- driver of speciation.
This may at first be surprising because
hybrid inviability appears as a consequence of recombining different, yet functionally equivalent, mechanisms.

Consistent with empirical observation of \emph{hybrid breakdown} \citep[e.g.][]{plotner2017chlorosis}), we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s,
since $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time relative to parentals, respectively.
\plr{refer to Turelli here}
% In this model, this is due to recombination shuffling system architectures, leaving the $F_2$ system homozygous for one parent at one particular locus and homozygous for the other parent at another possibly incompatible locus.
Another natural consequence of the model is Haldane's rule,
that if only one $F_1$ hybrid sex is inviable or sterile it is likely to be the heterogametic sex.
This occurs because if the genes underlying a regulatory network are distributed among both autosomes and the sex chromosome, 
then heterogametic $F_1$s show variation similar to that seen in $F_2$s.

@@

\paragraph{Nonlinearity}
Of course, real regulatory networks are not linear dynamical systems.
Most notably, physiological limits put upper bounds on expression levels,
implying saturating response curves \citep{michalis_mentin}.
It remains to be seen how well these results carry over into real systems,
but the fact that most nonlinear systems can be locally approximated by a linear one
suggests qualitatively similar results.
Furthermore, nonidentifiability (which implies the existence of neutral directions)
is often found in practice in moderately complex models of biological systems. \plr{need citation}

\paragraph{Peter: Quant gen discussion}
Maybe compare to \citet{fierst}?

Do we talk about the $G$ matrix and mutational correlation somewhere?

Note that islands or bottlenecks leave large $\sigma_N$ but small $N_e$ so may go fast.

What is Ne?  In a metapopulation depends on migr rate: classic Turelli paper.

This has implications for genetic load also!!

hominids; neanderthal load.

\section*{Acknowledgements}
We would like to thank Sergey Nuzhdin, Stevan Arnold, Erik Lundgren, and Hossein Asgharian for valuable discussion.

>>>>

A complex molecular machinery translates an organism's genome into the characteristics on which natural selection acts,
the phenotype.
% Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. 
%Attaining a general understanding of the functioning and evolution of this molecular machinery
It is an overarching goal of many biological subdisciplines to attain a general understanding of the function and evolution of this molecular machinery. 
For example, 
there is a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks
\citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy, pavlicev2012model}. 
Mathematical models of both particular regulatory networks
and the evolution of such systems in general
can provide guidance where intuition fails,
and thus has the potential to discover general principles in the organization of biological systems 
as well as provide concrete numerical predictions \citep{servedio2014not}.

@@

Although nonidentifiability may frustrate the occasional engineer or scientist, viewed from another angle,
this concept can provide a starting point for thinking about externally equivalent systems
-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
These functional symmetries manifest in convergent and parallel evolution, 
as well as \emph{developmental system drift}: the observation that
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
 % \jss{should hare2008sepsid be here? It's sequence but not ``system'' drift}
% are these good here? {taylor2016diverse -- maybe not, matsui2015regulatory -- I think this relevant, stergachis2014conservation}.
%
%AND THESE?
  %% Over the last several years, different computational models have been applied to study reproductive incompatibility and speciation in models of gene regulation \citep{porter2002speciation, tulchinsky, palmer2009dynamics, tulchinsky2014hybrid, khatari2015simple}.
  %\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model. Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybrid incompatibility when allopatric populations employ divergent readout languages. This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene. Furthermore, they report reproductive isolation primarily following directional selection for a change in expression levels in each allopatric population; the evidence for reproductive isolation following balancing selection is much weaker. Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection. 
%Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective. Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially. Otherwise hybrids between two allopatric populations have fairly high fitnesses.

In this paper we outline a theoretical framework to study the evolution of biological systems, such as gene regulatory networks.
We study the evolution of an optimally adapted population subject to stabilizing selection for phenotype -- that is the population evolves under constant selective and environmental pressures.
Even if the phenotype remains stable over evolutionary time, the underlying mechanism might not remain so,
%Even when stabilizing selection preserves phenotype over evolutionary time, the underlying mechanism may change, 
as many distinct (and mutationally connected) molecular pathways can realize identical phenotypes.
%where the optimal phenotype remains constant over evolutionary time,
%so that despite strong stabilizing selection for phenotype,
%neutral drift in the underlying genotype is possible.
Applying results from system theory, we present
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes, and show that any biological system, in principal, can undergo system drift, as there is never a unique system architecture per phenotype.
%which gives concrete expectations for how any such
%system can, in principal, undergo system drift and rewire.
Even under stabilizing selection, a population can explore the set of all possible phenotypically equivalent gene networks.
Phenotypically equivalent gene networks are not necessarily compatible with one another, as such system drift may result in the reproductive incompatibility between populations isolated for a sufficiently long period of time, even in the absence of any sort of adaptative, selective, or environmental change. Finally, in some cases we estimate that reproductive incompatibility due to system drift can manifest on timescales on the order of $N_e$ generations.
%Consequentially, two populations isolated for a sufficiently long period of time, 
%will likely produce inviable hybrids, despite the absence of adaptation, directional selection, or environmental change.
% Under these conditions, speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.

@@

Above we synthesize concepts and tools from molecular and evolutionary biology with tools from control theory
to study the evolution of a mechanistic model of the molecular genotype-phenotype map under stabilizing selection.
This model allows us to analytically describe the processes discussed verbally as
phenogenetic drift \citep{weiss2000phenogenetic} and developmental system drift \citep{true2001developmental}.
In this context, the Kalman decomposition \citep{kalman1963mathematical} implies that nearly all systems are nonidentifiable,
and gives an analytical description of all phenotypically equivalent gene networks.
System nonidentifiability, rather than being a computational nuisance,
implies the existence of axes of genetic variation that are not constrained by selection.
The independent movement of separated populations along these axes
can lead to reduction in hybrid viability, and hence speciation,
at a speed that depends on effective population size and amount of genetic variation. 
In this model, at biologically reasonable parameter values,
system drift is a significant -- and often rapid -- driver of speciation.
This may at first be surprising because
hybrid inviability appears as a consequence of recombining different, yet functionally equivalent, mechanisms. 
While the genetic incompatibilities contributing to hybrid inviability described in this paper share similarities with those in the commonly discussed Dobzhansky-Muller model, they may in fact share more in common with what has been described as the ``pathway model'' \citep{lindtke2015genetic}.

Consistent with empirical observation of \emph{hybrid breakdown} \citep[e.g.][]{plotner2017chlorosis}), we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s,
since $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time relative to parentals, respectively.
%\plr{refer to Turelli here}
% In this model, this is due to recombination shuffling system architectures, leaving the $F_2$ system homozygous for one parent at one particular locus and homozygous for the other parent at another possibly incompatible locus.
Another natural consequence of the model is Haldane's rule,
that if only one $F_1$ hybrid sex is inviable or sterile it is likely to be the heterogametic sex.
This occurs because if the genes underlying a regulatory network are distributed among both autosomes and the sex chromosome, 
then heterogametic $F_1$s show variation similar to that seen in $F_2$s.

@@

\paragraph{Nonlinearity and model assumptions}
Of course, real regulatory networks are not linear dynamical systems.
Most notably, physiological limits put upper bounds on expression levels,
implying saturating response curves \citep{michalis_mentin}.
It remains to be seen how well these results carry over into real systems,
but the fact that most nonlinear systems can be locally approximated by a linear one
suggests qualitatively similar results.
Furthermore, nonidentifiability (which implies the existence of neutral directions)
is often found in practice in moderately complex models of biological systems. \plr{need citation}

Finally, despite our model's precise separation of phenotype and kryptotype, this relationship in nature may be far more complicated as aspects of the kryptotype may be less ``hidden'' than we currently assume. For instance, attributes excluded from the phenotype as modelled here, ignore the potential energy costs associated with execessively large (nonminimal) kryptotypes, as well as the relationship between a specific network architecture and mutational robustness, as well as many other presently neglected phenomena. Despite the model's simplifications and potentially nontrivial oversights, it may still roughly reflect and shed light upon speciation patterns as they occur in nature.

%\paragraph{Peter: Quant gen discussion}
%Maybe compare to \citet{fierst}?
%
%Do we talk about the $G$ matrix and mutational correlation somewhere?
%
%Note that islands or bottlenecks leave large $\sigma_N$ but small $N_e$ so may go fast.
%
%What is Ne?  In a metapopulation depends on migr rate: classic Turelli paper.
%
%This has implications for genetic load also!!
%
%hominids; neanderthal load.
%
\section*{Acknowledgements}
We would like to thank Sergey Nuzhdin, Stevan Arnold, Erik Lundgren, and Hossein Asgharian for valuable discussion.
